### 2015-03-31 doc/document.tex
Uploaded paper draft.

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}

\begin{document}
\firstpage{1}

\title[short Title]{Seeq: a library for inexact DNA sequence matching}
\author[Zorita \textit{et~al}]{Eduard Zorita\,$^{1,2}$ and Guillaume J. Filion\,$^{1,2}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and
  Cancer Programme, Centre for Genomic Regulation (CRG), Dr. Aiguader
  88, 08003 Barcelona, Spain\\
$^{2}$Universitat Pompeu Fabra (UPF), 08002 Barcelona, Spain}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}

\section{Motivation:}
Searching specific sequences over DNA reads is an elemental procedure
of biological data analysis. Experimental DNA samples have usually
been engineered to contain pattern sequences at the positions where
operations such as adaptor removal, sequence trimming or barcode
extraction have to be performed. The experimental data contain errors
that must be tolerated in order to identify these patterns at their
correct positions. This process is straightforward in data produced by
second generation sequencing machines, in which the error rates lie
below 2\%. Short sequences with small error tolerance suffice to
identify the patterns accurately in these circumstances. However,
third generation sequencing technologies promise longer reads with
error rates around 15\%. Consequently, this opens a new scenario in
which both the pattern length and the relative error tolerance will
have to be dramatically increased.
\section{Results:}
In this paper we present seeq, an inexact pattern matching algorithm
optimized for DNA sequence matching with almost constant running
time. We benchmark seeq against the most widely used algorithms for
inexact pattern matching. The results show that seeq scales better and
becomes significantly faster as the error rate increases.
\section{Availability:}
Seeq is available in three forms: Linux software, C Library and Python
module. The source code and the C/Python libraries are available for
download at
\href{http://github.com/ezorita/seeq}{http://github.com/ezorita/seeq}.
\section{Contact:} \href{guillaume.filion@gmail.com}{guillaume.filion@gmail.com}
\end{abstract}

\section{Introduction}

Inexact string matching is an old problem that dates back to the
sixties with the definition of the edit distance \citep{Lev65}. The
first algorithms used to solve this problem were based on dynamic
programming \citep{NW70}. Although the algorithms were not very
efficient, they are still widely used to align biological sequences
due to their simplicity and flexibility. Later on, more sophisticated
algorithms derived from the Boyer-Moore search method \citep{BM77}
were designed to search patterns on long strings allowing
differences. Such algorithms are much faster than the dynamic
programming approach because they do not compute all the relative
distances. Instead, the pattern is pre-processed to design a search
strategy which skips parts of the text that do not contain any
match. The jumping strategy is very efficient for perfect matches or
small edit distances because the non-matching condition is inferred
with only a few comparisons. In this scenario, the number of
operations per input character is ridiculously small given that large
areas of the text are never processed. However, other scenarios are
not so favorable for this technique. For instance, if the proposed
pattern matches in almost any text position, the performance degrades
severely since the algorithm will not be able to jump so
often. Besides, for high mismatch ratios the search strategy becomes
complex and requires more comparisons to determine a non-matching
condition, specially when the alphabet is as small as in the task of
matching DNA sequences.

Sequencing data produced by the Illumina platform \citep{Mar05}
typically has 1-2\% error rate mainly consisting of substitutions
\citep{Doh08}. Therefore one expects to find all the matching strings
within a small edit distance relative to the original pattern
length. Moreover, if such patterns are produced artificially
(e.g. sequencing adaptors or barcodes), it is reliable to use short 
sequences that are not present in the target genome. On the other
side, the PacBio platform produces reads two orders of magnitude
longer than Illumina, but at the expense of a much higher error rate
(15\%) consisting principally of insertions and deletions
\citep{Eid09}. This high error rate dramatically increases the false
positive rate unless the pattern sequence is extended
accordingly. This brings us to a new scenario with unusually long
patterns and edit distances that lie on the same order of magnitude.

In this article we present \emph{seeq}, an algorithm based on a
deterministic finite automaton (DFA) structure. The automaton is built
as the input text is processed and has been optimized for DNA sequence
matching. We benchmark seeq against other widely used inexact string
matching software. The results show that seeq scales better and
becomes significantly faster as the error rate increases.

\begin{methods}
\section{Methods}

The algorithm proposed is the DFA representation of a dynamic
programming algorithm. The goal is to find the occurrences of a
pattern $P$ in a text $T$ that are $k$-differences or less. The
lengths of $P$ and $T$ are $m$ and $n$, respectively. The score matrix
$\mathbf{C} \in \mathbb{N}^{m\times n}$ is initialized so as that any
text position is a potential alignment start, i.e. $\mathbf{C}(0,j)=0$
and $\mathbf{C}(i,0) = i$. After the initialization, the score matrix
is computed and filled column-wise. Note that the standard dynamic
programming update process can be represented as
$\mathbf{c}_j=f(\mathbf{c}_{j-1},T_j)$. In effect, the value of the 
$j$-th column $\mathbf{c}_{j}$ is a deterministic transition that only
depends on $\mathbf{c}_{j-1}$ and the character $T_j$. We take
advantage of this property to build a deterministic automata in which
the different states represent the set of non-equivalent alignment
columns and the transitions between states are dictated by the input
characters.

To make the alignment algorithm efficient, only the significant
($\mathbf{C}(i,j) \le k$) parts of the column are computed. The DFA
states $S$ have a vector representation
$\mathbf{s} = \mbox{dif}(\mathbf{c}_j)$ generated by differentially
encoding the $j$-th column values as 
$\mathbf{s}(i)=\mathbf{C}(i+1,j)-\mathbf{C}(i,j)+1$ for
$i=0\ldots m-1$. This is done for two reasons: (i) the alphabet of
$\mathbf{s}$ is small \{0,1,2\}, therefore the alignment columns can be
efficiently stored in a ternary trie, and (ii) the inverse process to
recompute the original column ($\mathbf{c}_j=\mbox{undif}(\mathbf{s})$)
is performed without further information, provided that $\mathbf{C}(0,j)=0$.

\begin{figure}[!tpb]
%\centerline{\includegraphics[scale=0.5]{fig_methods.pdf}}
\caption{}\label{matrix}
\end{figure}

The DFA construction process is illustrated in Figure. X. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{itemize}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then $S = S_q$.
\item Otherwise, the vector $\mathbf{s}_j$ is computed as
$\mathbf{s}_j=\mbox{dif}(f(\mbox{undif}(\mathbf{s}), T_j))$.
If a state represented by $\mathbf{s}_j$ already exists ($S_q$), $S$
is connected to such state with a path labeled $T_j$ and $S = S_q$. On
the other hand, if $\mathbf{s}_j$ does not exist, a new DFA state
$S_j$ is created and both nodes are connected in the same
way, then $S = S_j$.
\end{itemize}
After several steps, the DFA graph is densely connected and the
complexity of the algorithm boils down to direct DFA state
transitions.

\end{methods}

\section{Results}

We have benchmarked seeq against two competitive inexact string
matching algorithms: agrep \citep{Wu92} and nr-grep \citep{Nav01}. We
used two real sequencing datasets for benchmarking. Both datasets contain
10 million reads but differ in read length. Dataset 1 consists random
Drosophila genome reads of length 100. Dataset 2 reads are 50
nucleotides long, from which the first 30 nucleotides are constant and
the last 20 are random (barcodes). The software were set to match
random patterns ranging from 5 to 50 nucleotides with a random error
tolerance up to 40\%. We would like to point out that agrep limits the
number of differences to 8, therefore the comparison shown here is
only fair between seeq and nr-grep.

Figure \ref{fig:results} shows the running time of each execution as
a function of the relative error tolerace ($k/m$). For low error
rates, both agrep and nr-grep perform better than seeq, with a maximum
5-fold advantage when $k=1$ and $m > 25$. On the other hand, seeq is
up to two orders of magnitude faster than nr-grep and one order of
magnitude faster than agrep for $k/m > 0.15$. The similarity between
the two datasets indicates that neither software is particularly
favored when the processed data has a repetitive structure. The worst
to best case ratio also speaks strongly in favor of seeq (5), which
achieves almost constant running time. This value is one and almost
three orders of magnitude higher for agrep (50) and nr-grep (1000),
respectively.

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.5]{results.pdf}}
\caption{Benchmark between seeq, agrep and nr-grep.}
\label{fig:results}
\end{figure}

\section{Conclusion}

In this article we have introduced and benchmarked seeq, an inexact
DNA sequence matching algorithm. We have measured and compared the
running time of seeq and two competitive software in different
configurations. The results conclude that seeq is faster in higher
error regions with a worst case speedup of almost 100. We attribute
the higher efficiency of seeq to the fact that it has been
specifically designed to match DNA sequences. Thus, we could take
advantage of algorithms that are not feasible in general string
matchers because they only scale well when the size of the alphabet is
reduced.

\paragraph{Funding\textcolon}

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{document}

\end{document}

### 2015-03-31 src/seeq.c
Enabled inverse mode again.

@@ -110,9 +110,7 @@ seeq
       int retval;
       while ((retval = seeqFileMatch(sqfile, sq, match_options)) > 0) {
          if (args.compact) fprintf(stdout, "%ld:%d-%d:%d",sq->match.line, sq->match.start, sq->match.end-1, sq->match.dist);
-         else if (args.invert) {
-            fprintf(stdout, "%s", sq->match.string);
-         } else {
+         else {
             if (args.showline) fprintf(stdout, "%ld ", sq->match.line);
             if (args.showpos)  fprintf(stdout, "%d-%d ", sq->match.start, sq->match.end-1);
             if (args.showdist) fprintf(stdout, "%d ", sq->match.dist);
@@ -140,8 +138,10 @@ seeq
                   fprintf(stdout, "%s", sq->match.string + sq->match.end);
                }
                else fprintf(stdout, "%s", sq->match.string);
+            } else if (args.invert) {
+               fprintf(stdout, "%s", sq->match.string);
             }
-         }
+      } 
          fprintf(stdout, "\n");
       }
       if (retval == -1) {

### 2015-03-31 doc/document.tex
Enabled inverse mode again.
First edits on the text.

@@
 \section{Motivation:}
-Searching specific sequences over DNA reads is an elemental procedure
-of biological data analysis. Experimental DNA samples have usually
-been engineered to contain pattern sequences at the positions where
-operations such as adaptor removal, sequence trimming or barcode
-extraction have to be performed. The experimental data contain errors
-that must be tolerated in order to identify these patterns at their
-correct positions. This process is straightforward in data produced by
-second generation sequencing machines, in which the error rates lie
-below 2\%. Short sequences with small error tolerance suffice to
-identify the patterns accurately in these circumstances. However,
-third generation sequencing technologies promise longer reads with
-error rates around 15\%. Consequently, this opens a new scenario in
-which both the pattern length and the relative error tolerance will
-have to be dramatically increased.
+Searching specific sequences in DNA is one of the most common tasks
+in the analysis of biological data. Experimental samples often contain
+known sequences at the positions where adaptor removal, sequence
+trimming or barcode extraction have to be performed. Experimental data
+contain errors that must be tolerated in order to identify these
+patterns. For reads produced by second generation sequencing,
+where the error rates lie below 2\%, short sequences with small error
+tolerance suffice to identify the patterns. However, third generation
+sequencing technologies promise longer reads at the cost of 10-fold
+higher error rates. In this scenario, both the pattern length and the
+relative error tolerance have to be dramatically increased.
 \section{Results:}
-In this paper we present seeq, an inexact pattern matching algorithm
-optimized for DNA sequence matching with almost constant running
-time. We benchmark seeq against the most widely used algorithms for
-inexact pattern matching. The results show that seeq scales better and
-becomes significantly faster as the error rate increases.
+Here we present seeq, an algorithm for pattern matching with errors.
+Seeq is optimized for DNA sequences and shows very little increase in
+running time as error tolerance increases. We benchmark seeq against
+the most widely used algorithms for inexact pattern matching. The
+results show that seeq scales better and becomes significantly faster
+as the error rate increases.
 \section{Availability:}
-Seeq is available in three forms: Linux software, C Library and Python
-module. The source code and the C/Python libraries are available for
-download at
+Seeq is available in three forms: a Linux software, a C library and a 
+Python module. The source code and the C/Python libraries are available
+for download at
 \href{http://github.com/ezorita/seeq}{http://github.com/ezorita/seeq}.
@@
 
-Inexact string matching is an old problem that dates back to the
-sixties with the definition of the edit distance \citep{Lev65}. The
-first algorithms used to solve this problem were based on dynamic
-programming \citep{NW70}. Although the algorithms were not very
-efficient, they are still widely used to align biological sequences
-due to their simplicity and flexibility. Later on, more sophisticated
-algorithms derived from the Boyer-Moore search method \citep{BM77}
-were designed to search patterns on long strings allowing
-differences. Such algorithms are much faster than the dynamic
+Inexact string matching dates back to the origin of computer sciences
+\citep{Lev65}. The first algorithms addressing this problem were based
+on dynamic programming \citep{NW70}. They are still widely used to
+align biological sequences due to their simplicity and flexibility,
+More sophisticated algorithms derived from the Boyer-Moore search
+method \citep{BM77} were later designed to search patterns on long
+strings allowing errors. Such algorithms are faster than the dynamic
 programming approach because they do not compute all the relative
@@
 match. The jumping strategy is very efficient for perfect matches or
-small edit distances because the non-matching condition is inferred
-with only a few comparisons. In this scenario, the number of
-operations per input character is ridiculously small given that large
-areas of the text are never processed. However, other scenarios are
-not so favorable for this technique. For instance, if the proposed
-pattern matches in almost any text position, the performance degrades
-severely since the algorithm will not be able to jump so
-often. Besides, for high mismatch ratios the search strategy becomes
-complex and requires more comparisons to determine a non-matching
-condition, specially when the alphabet is as small as in the task of
-matching DNA sequences.
+small edit distances because non-matching conditions can be inferred
+with only a few comparisons. In those conditions, the number of
+operations per input character is small because large areas of the
+text are never processed. However, when the error rate is high, the
+search strategy requires more comparisons to identify a non-matching
+condition. The drop in performance is even more accute when the alphabet
+is as small as in the case of DNA sequences.
 
-Sequencing data produced by the Illumina platform \citep{Mar05}
-typically has 1-2\% error rate mainly consisting of substitutions
-\citep{Doh08}. Therefore one expects to find all the matching strings
-within a small edit distance relative to the original pattern
-length. Moreover, if such patterns are produced artificially
-(e.g. sequencing adaptors or barcodes), it is reliable to use short 
-sequences that are not present in the target genome. On the other
-side, the PacBio platform produces reads two orders of magnitude
-longer than Illumina, but at the expense of a much higher error rate
-(15\%) consisting principally of insertions and deletions
-\citep{Eid09}. This high error rate dramatically increases the false
+Illumina reads typically have a 1-2\% error rate \citep{Mar05},
+mainly consisting of substitutions \citep{Doh08}, which are ideal
+conditions for the Boyer-Moore algorithm. However, third generation
+sequencing technologies such as the PacBio platform now come of age.
+They produce reads two orders of magnitude longer than Illumina, at
+the expense of a 10-fold higher error rate (15\%) consisting mostly
+of insertions and deletions \citep{Eid09}.  
+This high error rate dramatically increases the false
 positive rate unless the pattern sequence is extended
-accordingly. This brings us to a new scenario with unusually long
-patterns and edit distances that lie on the same order of magnitude.
+accordingly. This raises a new scenario with long patterns and
+large edit distances.
 
-In this article we present \emph{seeq}, an algorithm based on a
-deterministic finite automaton (DFA) structure. The automaton is built
-as the input text is processed and has been optimized for DNA sequence
-matching. We benchmark seeq against other widely used inexact string
-matching software. The results show that seeq scales better and
+Here we present \emph{seeq}, an algorithm using a dynamic version of
+deterministic finite automata (DFA). Instead of using standard
+construction procedures, the dynamic DFA of seeq is built as the input
+text is processed. We benchmark seeq against other widely used inexact
+string matching software. The results show that seeq scales better and
 becomes significantly faster as the error rate increases.
@@
 
-The algorithm proposed is the DFA representation of a dynamic
-programming algorithm. The goal is to find the occurrences of a
-pattern $P$ in a text $T$ that are $k$-differences or less. The
-lengths of $P$ and $T$ are $m$ and $n$, respectively. The score matrix
-$\mathbf{C} \in \mathbb{N}^{m\times n}$ is initialized so as that any
-text position is a potential alignment start, i.e. $\mathbf{C}(0,j)=0$
-and $\mathbf{C}(i,0) = i$. After the initialization, the score matrix
-is computed and filled column-wise. Note that the standard dynamic
-programming update process can be represented as
-$\mathbf{c}_j=f(\mathbf{c}_{j-1},T_j)$. In effect, the value of the 
-$j$-th column $\mathbf{c}_{j}$ is a deterministic transition that only
-depends on $\mathbf{c}_{j-1}$ and the character $T_j$. We take
-advantage of this property to build a deterministic automata in which
-the different states represent the set of non-equivalent alignment
-columns and the transitions between states are dictated by the input
-characters.
-
-To make the alignment algorithm efficient, only the significant
-($\mathbf{C}(i,j) \le k$) parts of the column are computed. The DFA
-states $S$ have a vector representation
-$\mathbf{s} = \mbox{dif}(\mathbf{c}_j)$ generated by differentially
-encoding the $j$-th column values as 
-$\mathbf{s}(i)=\mathbf{C}(i+1,j)-\mathbf{C}(i,j)+1$ for
-$i=0\ldots m-1$. This is done for two reasons: (i) the alphabet of
-$\mathbf{s}$ is small \{0,1,2\}, therefore the alignment columns can be
-efficiently stored in a ternary trie, and (ii) the inverse process to
-recompute the original column ($\mathbf{c}_j=\mbox{undif}(\mathbf{s})$)
-is performed without further information, provided that $\mathbf{C}(0,j)=0$.
+We use a DFA representation of a dynamic programming algorithm.
+We search the occurrences of the pattern $P$ of length $m$
+in a text $T$ of length $n$, with up to $k$ differences.
+The edit matrix $\mathbf{C} \in \mathbb{N}^{m\times n}$ is initialized
+with $\mathbf{C}(0,j)=0$ and $\mathbf{C}(i,0) = i$, and then updated
+column-wise with the Needleman-Wunsch recusrion. Note that this process
+can be represented as $\mathbf{c}_j=f(\mathbf{c}_{j-1},T_j)$,
+\textit{i.e.} the value of the $j$-th column $\mathbf{c}_{j}$ is a
+deterministic transition that only depends on $\mathbf{c}_{j-1}$ and the
+character $T_j$. We take advantage of this property to build a
+deterministic automaton where the states are the columns of the matrix
+and the transitions are dictated by the input characters.
+We represent the states of the DFA by the differential encoding
+of the $j$-th column, $\mathbf{s}(i)=\mathbf{C}(i+1,j)-\mathbf{C}(i,j)+1$
+so that the state can be stored in a ternary trie ($\mathbf{s}(i)$
+can take only 3 values).
 
@@
 the active DFA state, the algorithm recursively proceeds as follows:
-\begin{itemize}
+\begin{enumerate}
 \item If $S$ has an outgoing path (to $S_q$) with
-label $T_j$, then $S = S_q$.
-\item Otherwise, the vector $\mathbf{s}_j$ is computed as
-$\mathbf{s}_j=\mbox{dif}(f(\mbox{undif}(\mathbf{s}), T_j))$.
-If a state represented by $\mathbf{s}_j$ already exists ($S_q$), $S$
-is connected to such state with a path labeled $T_j$ and $S = S_q$. On
-the other hand, if $\mathbf{s}_j$ does not exist, a new DFA state
-$S_j$ is created and both nodes are connected in the same
-way, then $S = S_j$.
-\end{itemize}
-After several steps, the DFA graph is densely connected and the
-complexity of the algorithm boils down to direct DFA state
+label $T_j$, then set $S = S_q$.
+\item Otherwise, compute the column of the edit matrix corresponding
+to $f(S, T_j)$. Case 2a: if this state already exists ($S_q$), connect
+$S$ to it with a path labeled $T_j$ and set $S = S_q$. Case 2b: if
+the state does not exist, create it, and do as in in the previous
+case 2a.
+\end{enumerate}
+
+After several steps, the graph of the DFA is densely connected and
+the steps of state and edge creation are almost never performed. 
+The complexity of the algorithm is then reduced to direct DFA state
 transitions.
@@
 
-We have benchmarked seeq against two competitive inexact string
+We have benchmarked seeq against two standard inexact string
 matching algorithms: agrep \citep{Wu92} and nr-grep \citep{Nav01}. We
-used two real sequencing datasets for benchmarking. Both datasets contain
-10 million reads but differ in read length. Dataset 1 consists random
-Drosophila genome reads of length 100. Dataset 2 reads are 50
-nucleotides long, from which the first 30 nucleotides are constant and
-the last 20 are random (barcodes). The software were set to match
+used two sequencing datasets for benchmarking. Both datasets contain
+10 million reads but differ in read length. Dataset 1 consists of random
+\textit{Drosophila} genome reads of length 100. The reads of dataset 2
+are 50 nucleotides long, from which the first 30 nucleotides are constant
+and the last 20 are random (barcodes). The software were set to match
 random patterns ranging from 5 to 50 nucleotides with a random error
-tolerance up to 40\%. We would like to point out that agrep limits the
-number of differences to 8, therefore the comparison shown here is
-only fair between seeq and nr-grep.
+tolerance up to 40\%. Importantly, agrep limits the number of differences
+to 8, which is a serious limitation and makes the comparison fair only between seeq and nr-grep.
 
-Figure \ref{fig:results} shows the running time of each execution as
-a function of the relative error tolerace ($k/m$). For low error
+The running time of each execution is plotted against the relative
+error tolerace ($k/m$) on figure \ref{fig:results}. For low error
 rates, both agrep and nr-grep perform better than seeq, with a maximum
@@
 magnitude faster than agrep for $k/m > 0.15$. The similarity between
-the two datasets indicates that neither software is particularly
-favored when the processed data has a repetitive structure. The worst
-to best case ratio also speaks strongly in favor of seeq (5), which
-achieves almost constant running time. This value is one and almost
-three orders of magnitude higher for agrep (50) and nr-grep (1000),
-respectively.
+the two datasets indicates that neither software is favored when the
+data has a repetitive structure. Also note that seeq 
+achieves near-constant running time.
 
@@
 \centerline{\includegraphics[scale=0.5]{results.pdf}}
-\caption{Benchmark between seeq, agrep and nr-grep.}
+\caption{Benchmark between seeq, agrep and nr-grep. Each dot represents
+the running time of one search with a random pattern and a random
+edit distance. Note the logarithmic scale on the y axis.}
 \label{fig:results}
@@
 
-In this article we have introduced and benchmarked seeq, an inexact
-DNA sequence matching algorithm. We have measured and compared the
-running time of seeq and two competitive software in different
-configurations. The results conclude that seeq is faster in higher
-error regions with a worst case speedup of almost 100. We attribute
-the higher efficiency of seeq to the fact that it has been
-specifically designed to match DNA sequences. Thus, we could take
-advantage of algorithms that are not feasible in general string
-matchers because they only scale well when the size of the alphabet is
-reduced.
+The benchmark shows that seeq is faster in higher error regions with
+an excellent worst case. Actually, the running time of seeq depends
+very little on the error tolerance. The higher efficiency of seeq
+is attributed to the underlying dynamic DFA structure. Here, the key
+insight was to take advantage of the limited size of the DNA alphabet.
+This strategy is usually not efficient in general string matching
+problems.
+
+In practice, most pattern searches in DNA reads are performed from
+scripts and pipelines. To ease the use of seeq in existing software,
+we have ensured that it is comliant with Linux pipes. Also, we have
+created a Python module that can be imported and integrated seamlessly,
+as it can be called on Python string objects that can be further
+manipulated. We also provide a API in C for the cases that optimal
+performance is required.
 

### 2015-04-01 doc/document.tex
Updated manuscript.

@@
 Here we present seeq, an algorithm for pattern matching with errors.
-Seeq is optimized for DNA sequences and shows very little increase in
-running time as error tolerance increases. We benchmark seeq against
+Seeq is optimized for DNA sequences and shows very little variance in
+running time with different configurations. We benchmark seeq against
 the most widely used algorithms for inexact pattern matching. The
@@
 on dynamic programming \citep{NW70}. They are still widely used to
-align biological sequences due to their simplicity and flexibility,
+align biological sequences due to their simplicity and flexibility.
 More sophisticated algorithms derived from the Boyer-Moore search
@@
 small edit distances because non-matching conditions can be inferred
-with only a few comparisons. In those conditions, the number of
+with only a few comparisons. In such conditions, the number of
 operations per input character is small because large areas of the
@@
 construction procedures, the dynamic DFA of seeq is built as the input
-text is processed. We benchmark seeq against other widely used inexact
-string matching software. The results show that seeq scales better and
+text is processed. We benchmark seeq against other inexact string
+matching software. The results show that seeq scales better and
 becomes significantly faster as the error rate increases. 
-
 \begin{methods}
@@
 \begin{figure}[!tpb]
-%\centerline{\includegraphics[scale=0.5]{fig_methods.pdf}}
-\caption{}\label{matrix}
+\centerline{\includegraphics[scale=0.6]{Figure_1.pdf}}
+\caption{Dynamic DFA construction process for T='TGTGCA' and P='GCAC'.
+a) Dynamic programming matrix of the alignment. The differential code
+of each column is shown inside the semicircles. b) State of the DFA at
+the end of the alignment.}\label{matrix}
 \end{figure}
 
-The DFA construction process is illustrated in Figure. X. The
+The DFA construction process is illustrated in Figure. \ref{matrix}. The
 dynamic programming algorithm starts at the DFA state
@@
 label $T_j$, then set $S = S_q$.
-\item Otherwise, compute the column of the edit matrix corresponding
-to $f(S, T_j)$. Case 2a: if this state already exists ($S_q$), connect
-$S$ to it with a path labeled $T_j$ and set $S = S_q$. Case 2b: if
-the state does not exist, create it, and do as in in the previous
-case 2a.
+\item Otherwise, revert $\mathbf{c}$ from $\mathbf{s}$ and compute
+the $j$-th column of the edit matrix as $\mathbf{c}_j = f(\mathbf{c}, T_j)$ and
+its diferential code $\mathbf{s}_j$. Then: a) If $\mathbf{s}_j$
+represents an existing state ($S_q$), connect $S$ to it with a path
+labeled $T_j$ and set $S = S_q$. b) If none of the states is
+represented by $\mathbf{s}_j$, create $S_j$, connect $S$ to it with a
+path labeled $T_j$ and set $S = S_j$.
 \end{enumerate}
@@
 tolerance up to 40\%. Importantly, agrep limits the number of differences
-to 8, which is a serious limitation and makes the comparison fair only between seeq and nr-grep.
+to 8, which is a serious limitation and makes the comparison fair only
+between seeq and nr-grep.
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[scale=0.5]{results.pdf}}
+\caption{Benchmark between seeq, agrep and nr-grep. Each dot represents
+the running time of one search with a random pattern and a random
+edit distance. Note the logarithmic scale on the y axis.}
+\label{fig:results}
+\end{figure}
 
@@
 achieves near-constant running time.
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[scale=0.5]{results.pdf}}
-\caption{Benchmark between seeq, agrep and nr-grep. Each dot represents
-the running time of one search with a random pattern and a random
-edit distance. Note the logarithmic scale on the y axis.}
-\label{fig:results}
-\end{figure}
 
@@
 as it can be called on Python string objects that can be further
-manipulated. We also provide a API in C for the cases that optimal
+manipulated. We also provide an API in C for the cases that optimal
 performance is required.

### 2015-04-01 doc/document.tex
Added funding and corrected typos. Ready to send.

@@
 
-Illumina reads typically have a 1-2\% error rate \citep{Mar05},
+Illumina reads typically have a 1-2\% error rate,
 mainly consisting of substitutions \citep{Doh08}, which are ideal
@@
 \centerline{\includegraphics[scale=0.6]{Figure_1.pdf}}
-\caption{Dynamic DFA construction process for T='TGTGCA' and P='GCAC'.
-a) Dynamic programming matrix of the alignment. The differential code
-of each column is shown inside the semicircles. b) State of the DFA at
+\caption{Dynamic DFA construction process. In this example, the search
+text is TGTGCA and the pattern is GCAC.
+A) Dynamic programming matrix of the alignment. The differential code
+of each column is shown inside the semicircles. B) State of the DFA at
 the end of the alignment.}\label{matrix}
@@
 
-The DFA construction process is illustrated in Figure. \ref{matrix}. The
+The DFA construction process is illustrated in Figure \ref{matrix}. The
 dynamic programming algorithm starts at the DFA state
@@
 After several steps, the graph of the DFA is densely connected and
-the steps of state and edge creation are almost never performed. 
-The complexity of the algorithm is then reduced to direct DFA state
-transitions.
+case 2 almost never occurs. The complexity of the algorithm is then
+reduced to direct DFA state transitions.
 
@@
 The running time of each execution is plotted against the relative
-error tolerace ($k/m$) on figure \ref{fig:results}. For low error
+error tolerace ($k/m$) on Figure \ref{fig:results}. For low error
 rates, both agrep and nr-grep perform better than seeq, with a maximum
@@
 scripts and pipelines. To ease the use of seeq in existing software,
-we have ensured that it is comliant with Linux pipes. Also, we have
+we have ensured that it is compliant with Linux pipes. Also, we have
 created a Python module that can be imported and integrated seamlessly,
@@
 \paragraph{Funding\textcolon}
+The research leading to these results has received funding from the
+European Research Council under the European Union's Seventh Framework
+Programme (FP7/2007-2013) / ERC SYNERGY n$^\circ$ 609989.
 

### 2015-04-02 doc/document.tex
Corrected typo in manuscript.

@@
 with $\mathbf{C}(0,j)=0$ and $\mathbf{C}(i,0) = i$, and then updated
-column-wise with the Needleman-Wunsch recusrion. Note that this process
+column-wise with the Needleman-Wunsch recursion. Note that this process
 can be represented as $\mathbf{c}_j=f(\mathbf{c}_{j-1},T_j)$,
@@
 \begin{figure}[!tpb]
-\centerline{\includegraphics[scale=0.6]{Figure_1.pdf}}
+\centerline{\includegraphics[scale=0.6]{Figure_1.eps}}
 \caption{Dynamic DFA construction process. In this example, the search
@@
 \begin{figure}[!tpb]
-\centerline{\includegraphics[scale=0.5]{results.pdf}}
+\centerline{\includegraphics[scale=0.5]{results.eps}}
 \caption{Benchmark between seeq, agrep and nr-grep. Each dot represents

### 2015-05-07 doc/response.tex
Wrote the first version of response to reviewers.

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{graphicx}

\title{A response to the review of the manuscript
``Seeq: a library for inexact DNA sequence matching''}
\date{}

\newenvironment{code}{\ttfamily}{\par}

\begin{document}
\maketitle
As the authors of the manuscript ``Seeq: a library for inexact
DNA sequence matching'', we would like to discuss some of the
ideas that were raised by the reviewers. We believe that
critical discussion, based on facts and arguments is fundamental
to the scientific process. We fully respect the decision of the
editor and of our peers, but we do not consider that we should
interrupt a fertile discussion because the manuscript was
rejected. Please do not see in this letter a criticism of the
reviews, but simply a wish to respond to some of the points that
were raised.

Because this is not a rebuttal, we will do our best to make it
informal and didactic.
But before doing so, we would like to congratulate the reviewers for
a really outstanding work. We are all for software quality and
excellent science, we wish that every tool were scrutinized with
the same amout of care. We have also learned important lessons,
and the suggestions will greatly help us improve the software and
the manuscript.

\section{The inception}

It started from casual Internet
browsing of pages that have little to do with bioinformatics.
The first is a seminal series of blog posts by Cox Russel [1]
explaining how and why he implemented the regular expression library
RE2. The key insight was that many modern regular expression engines
(including those of Perl and Python) have ``forgotten'' the
theory and are clearly dominated by lazy DFA-based implementations.

The second source of inspiration was a blog post by Nick Johnson [2]
where he explains the use of Levenshtein automata in the problem of
approximate string query. In Fig.~\ref{generalities} we give some
generalities about Levenshtein automata, in case the reader is not
familiar with these concepts.

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=.6]{edit_matrix.pdf}}
\caption{Generalities on Levenshtein automata. a) Edit graph
of the string \texttt{GA} with 1 tolerated error. The edit graph
is a NFA with state transition as shown in the top panel. States
are labelled by two numbers indicating their number of errors and
their position relative to the search pattern. NFAs can be ins
several active states (dark circles). The initial active states
of the NFA are as shown in the second panel. After reading
\texttt{A} from the input, the active states are updated as shown
on the third panel. The active states can be simplified by keeping
only the upper node on each column, as shown on the bottom panel.
b) DFA representation. Every NFA has a DFAs representation.
States of the DFA are sets of active states of the NFA. In a
lazy DFA, only the states required to process the input are present
(bottom panel).
}\label{generalities}
\end{figure}

When we had to identify spike oligos in our
sequencing experiments, we used the Python implementation suggested
by Nick Johnson, which solved the problem (Illumina sequencing is
not indel-prone, but oligos synthesis is, which is why we needed
the Levenshtein distance). However, the Python prototype was rather
slow, and because it is a non lazy DFA, the memory footprint was high.

We connected the ideas of the two posts, and the first C
implementation of Seeq gave astonishing results. We knew that this
was not novel (lazy DFAs were implemented in egrep since 1989), but
because Seeq was much faster then Myers' bitfield algorithm,
we thought that lazy Levenshtein automata did not quite
make the jump from string processing to bioinformatics and that
someone should tell the community that they are efficient.

\section{Sparsity of the DFA}
\label{sketch}

At that stage, we were both surprised and skeptical. One concern (also
pointed out by reviewer 2) was that the potential number of states
was very high. Our initial implementation was based on the NFA
called the edit graph (Fig.~\ref{generalities}a),
which can be in $2^{(k+1)L}$ states, where $k$
is the tolerated number of errors and $L$ is the length of the
pattern. This number is enormous, but surprisingly, the memory
footprint was never an issue,
even for large files and long patterns. For some reason,
the lazy Levenshtein automaton explored only a small portion of its
states.

We realized that we could reduce the possible number of states by
keeping only the upper envelope of the NFA states in the edit graph
(in the manuscript, we presented states as columns from the alignment
because we thought that bioinformaticians would be more comfortable
with this representation, but the representations are strictly
equivalent). This made Seeq faster and reduced the memory footprint.
And it also allows us to explain why the DFA is sparse.

With this representation, a state is a line of the edit graph that
extends from the top-left
corner. On this line, local maxima have travelled from the top-left
corner by going either to the right (in case of a match) or in
diagonal (in case of a mismatch). So the probability that a local
maximum is at position $(i,j)$, where $(0,0)$ is
the top-left corner is approximately

\begin{equation}
{j \choose i} \left(\frac{3}{4}\right)^i
  \left(\frac{1}{4}\right)^{j-i}.
\end{equation}

This expresses that the number of diagonal steps has a binomial
distribution with probability $3/4$. For fixed $j$, the average
depth of local maxima is $3j/4$. Using R, we can get the 95\%
percentiles of $i$ for the first 20 values of $j$. We obtain
\texttt{0  0  1  1  2  3  3  4  5  5  6  6  7  8  8  9 10 10 11 12}.
All the DFA states are lines above the diagonal, and approximately
95\% of them will be beneath the 95\% percentiles of $i$. A graphical
representation for a pattern of 20 nucleotides with up to 12 errors
gives the idea that the majority of the states are squeezed within
a narrow region of the edit graph (Fig.~\ref{sparsity}).

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.24]{sparsity_of_the_DFA.pdf}}
\caption{The Levenshtein automaton of Seeq is sparse. The sketch
represents the edit graph for a pattern of length 20 with 12
tolerated errors. The dark dots cover a region where active
states are approximately 90-95\% of the time. This region is
substantially smaller than the full edit graph and contains
221,930 different DFA states.
}\label{sparsity}
\end{figure}

It is not trivial to compute the number of states in this region
in general terms, however it can be done for the example above.
There are 221,930 states \textit{i.e.} only $0.006\%$ of the
$3.5 \cdot 10^9$ upper bound computed from the formula provided by
reviewer 2. This is of course a heuristic argument that does not
resist worst case analysis. But the upshot is that on average,
Seeq needs to store very few states of the Levenshtein automaton.

\section{Performance}

The sketch of analysis above explains all the properties of Seeq.
Because the number of states is much lower than expected, Seeq
rarely needs to use dynamic programming to compute a new state.
As an added benefit, there will be few cache misses because
most of the time, Seeq only references the same small
set of sates.

Let us give a few examples to highlight the behavior of the
algorithm. When searching the 20 nucleotide pattern
\texttt{GATGTAGCGCGATTAGCCTG}
in hg19 with up to 4 errors, Seeq ended with 5,294 states and spent
0.01\% of the time in dynamic programming. Even though
it had to scan the whole genome, the run was executed in 27 seconds,
compared to 54 seconds for Linux \texttt{wc}, which only counts the
number of characters. On such problems where the number of operations
per character is small, the function used to read the file makes
the difference. We used GNU's \texttt{readline} because it was
faster than the alternatives in our initial tests.

Allowing 12 errors, which is the case of the sketch
of proof given in section \ref{sketch}, Seeq ended with 170,679 states,
close to our approximation. In this case, the running time
was 1 minute 18 seconds, of which 0.39 seconds were spent in dynamic
programming. So the difference with the previous case is either
due to cache misses or to the additional time required to process
the hits (of which there is a lot in this example).

What about longer patterns with higher error tolerance?
When searching the 42 nucleotide pattern
\texttt{GATGTAGCGCGATTAGCCTGAAAATGCGAGTACGGCGCGAAT} with
up to 12 errors, Seeq ran for 2 minutes 11 seconds and ended with
3,486,701 states. About 20 seconds were spent in dynamic programming,
which shows that it is not the bottleneck.
Even with very efficient vectorization to speed up this step, the
performance of Seeq would not improve substantially. The increase
in running time is more likely due to the increased number of cache
misses.

In comparison, Myers' bitfield algorithm ran for 66 minutes
44 seconds on this problem. Counting 17 operations per character,
this is about 75 ns per operation, which is the right order of
magnitude. For the bitfield algorithm to beat Seeq on the last problem,
each operation would have to be processed in less than 2.3 ns. We
were very surprised by this result, because Myers' bitfield
algorithm has the reputation of being the fastest known alignment
algorithm. But it needs to do 17 operations per character, whereas
most of the time Seeq does just 1, and it is often a cache hit. On
the last problem, Seeq performs 1 operation on 99.7\% of the
nucleotides. In summary, Myers' bitfield is not the fastest
alignment algorithm, it is the fastest alignment algorithm with
no memory footprint.

If lazy Levenshtein automata are so good, why were they not used
widely before? This is a good question. We surmise that this has
to do with the evolution of hardware. At the time they were proposed
in bioinformatics, \textit{i.e.} at the end of the 1990's, computers
typically had around 100 Mb of RAM, so they were probably considered
memory hungry and other methods filled the niche.

It is important to note that in the last case, the memory footprint
peaked at 1.5 Gb, which is approximately 430 bytes per state. The
states of the DFA proper are only an array of 5 pointers (40 bytes),
but Seeq indexes them in a ternary tree, which represents the bulk
of the memory usage. So it is true that the number of states will
become intractable for very large patterns and very high number of
tolerated errors.

\section{Concluding remarks}

We still believe that Seeq has an interesting message for the
bioinformatics community. It is certainly not the right tool for
every job, it is not intended to replace BLAST or BWA. Yet there
is a sweet spot for pattens around 20-30 nucleotides where we
reach good performance even for a hig number of tolerated errors.

Most sequences of this size are unique, even in large genomes,
which is why spike-ins, watermarks, adapters, barcodes and
other artificial sequences that need to be identified in sequencing
experiments will be in this range. The target application of 
Seeq is thus the Python library because it allows to easily write
parsers to identify the pattern and further process the reads as
simple Python strings.

We are aware that the information provided here would have made
the manuscript much stronger, but with a two page limit we had to
remove important parts. The difficulty was that Seeq is not novel
enough to deserve a full feature article, and it is too complex
to be described in just two pages.
We decided to present Seeq from the
perspective of the user, explain the principle and show the
performance. To be honest, we thought that there would be a
round of revisions where we could explain all this and decide
together with the reviewers how much should go in the article.
We realize now that we did not do a very good job at explaining
why Seeq is interesting. But we learned our lessons, we will do
better next time.

\section{References}

[1] https://swtch.com/~rsc/regexp/regexp1.html
\par\noindent
[2] http://blog.notdot.net/2010/07/Damn-Cool-Algorithms-Levenshtein-Automata

\end{document}

### 2015-05-07 setup.py

@@ -13,7 +13,5 @@ setup (name = 'seeq',
        author = 'Eduard V. Zorita',
        author_email = 'eduardvalera@gmail.com',
        url = 'https://docs.python.org/extending/building',
-       long_description = '''
-Seeq library for Python
-''',
+       long_description = 'Seeq library for Python\n',
        ext_modules = [module1])

### 2015-05-07 src/libseeq.c

@@ -61,8 +61,8 @@ seeqNew
 //
 // SIDE EFFECTS:
 //   The returned seeq_t structure must be freed using 'seeqFree'.
-
 {
+
    // Check parameters.
    if (mismatches < 0) {
       seeqerr = 1;
@@ -921,6 +921,7 @@ dfa_newstate
 //   The dfa may be reallocated, so the content of *dfap may be different
 //   at the end of the funcion.
 {
+
    // Set error to 0.
    seeqerr = 0;
 

### 2015-05-08 doc/myers_bit_vector.c
Improved type consistency. Added 'make dev' for development compiling.

@@ -1,0 +1,87 @@
+#define _GNU_SOURCE
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+
+typedef uint64_t bitV_t;
+
+static const char SIGMA[4] = "ACGT";
+static const int TRANSLATE[256] = {
+   ['a'] = 1, ['c'] = 2, ['g'] = 3, ['t'] = 4,
+   ['A'] = 1, ['C'] = 2, ['G'] = 3, ['T'] = 4,
+};
+
+int
+search
+(
+   const int m,
+   bitV_t Peq[],
+   char *txt,
+   int k
+)
+{
+   bitV_t Pv = ~0;
+   bitV_t Mv =  0;
+   int score = m;
+
+   for (int j = 0 ; txt[j] != 0 ; j++) {
+      bitV_t Eq = Peq[txt[j]];
+      bitV_t Xv = Eq | Mv;
+      bitV_t Xh = (((Eq & Pv) + Pv) ^ Pv) | Eq;
+      bitV_t Ph = Mv | ~ (Xh | Pv);
+      bitV_t Mh = Pv & Xh;
+
+      if (Ph & (1 << m-1)) {
+         score++;
+      }
+      else if (Mh & (1 << m-1)) {
+         score--;
+      }
+
+      Ph = Ph << 1;
+      Mh = Mh << 1;
+      Pv = Mh | ~ (Xv | Ph);
+      Mv = Ph & Xv;
+
+      if (score < k+1) {
+         fprintf(stdout, "%d\n", j);
+      }
+   }
+
+}
+
+int main(int argc, char *argv[]) {
+
+   bitV_t Peq[5] = {0,0,0,0,0};
+   char query[] = "GATGTAGCGCGATTAGCCTGAAAATGCGAGTACGGCGCGAAT";
+   int m = strlen(query);
+   // Pre-compute 'Peq'.
+   for (int s = 0 ; s < strlen(SIGMA) ; s++) {
+      char t = SIGMA[s];
+      for (int i = 0 ; i < m ; i++) {
+         if (query[i] == t) Peq[s+1] = Peq[s+1] | (1 << i);
+      }
+   }
+
+   char *seq = NULL;
+   ssize_t nread;
+   size_t buffer_size = 256; 
+   char *line = malloc(buffer_size * sizeof(char));
+
+   FILE *inputf = fopen(argv[1], "r");
+   if (inputf == NULL) abort();
+
[…]

### 2015-05-08 src/libseeq.c

@@ -27,17 +27,17 @@
 
 int seeqerr = 0;
 
-char * seeq_strerror[11] = {   "Check errno",
-                               "Illegal matching distance value",
-                               "Incorrect pattern (double opening brackets)",
-                               "Incorrect pattern (double closing brackets)",
-                               "Incorrect pattern (illegal character)",
-                               "Incorrect pattern (missing closing bracket)",
-                               "Illegal path value passed to 'trie_search'",
-                               "Illegal nodeid passed to 'trie_getrow' (specified node is not a leaf).\n",
-                               "Illegal path value passed to 'trie_insert'",
-                               "Pattern length must be larger than matching distance",
-                               "Passed seeq_t struct does not contain a valid file pointer"};
+static const char * seeq_strerror[11] = {"Check errno",
+                                  "Illegal matching distance value",
+                                  "Incorrect pattern (double opening brackets)",
+                                  "Incorrect pattern (double closing brackets)",
+                                  "Incorrect pattern (illegal character)",
+                                  "Incorrect pattern (missing closing bracket)",
+                                  "Illegal path value passed to 'trie_search'",
+                                  "Illegal nodeid passed to 'trie_getrow' (specified node is not a leaf).\n",
+                                  "Illegal path value passed to 'trie_insert'",
+                                  "Pattern length must be larger than matching distance",
+                                  "Passed seeq_t struct does not contain a valid file pointer"};
 
 seeq_t *
 seeqNew
@@ -122,8 +122,8 @@ seeqNew
    // Initialize match_t struct.
    sq->match.start  = 0;
    sq->match.end    = 0;
-   sq->match.dist   = mismatches + 1;
    sq->match.line   = 0;
+   sq->match.dist   = mismatches + 1;
    sq->match.string = NULL;
 
    return sq;
@@ -151,8 +151,8 @@ seeqFree
[…]

### 2015-05-08 src/libseeq.h

@@ -50,11 +50,11 @@ typedef struct match_t    match_t;
 typedef struct seeqfile_t seeqfile_t;
 
 struct match_t {
-   int    start;
-   int    end;
-   int    dist;
-   long   line;
-   char * string;
+   size_t   start;
+   size_t   end;
+   size_t   line;
+   int      dist;
+   char   * string;
 };
 
 struct seeq_t {
@@ -68,22 +68,22 @@ struct seeq_t {
 };
 
 struct seeqfile_t {
-   long    line;
+   size_t  line;
    FILE  * fdi;
 };
 
-seeq_t *    seeqNew         (const char *, int);
-void        seeqFree        (seeq_t *);
-seeqfile_t* seeqOpen        (const char *);
-int         seeqClose       (seeqfile_t *);
-long        seeqFileMatch   (seeqfile_t *, seeq_t *, int);
-long        seeqStringMatch (const char *, seeq_t *, int);
-long        seeqGetLine     (seeq_t *);
-int         seeqGetDistance (seeq_t *);
-int         seeqGetStart    (seeq_t *);
-int         seeqGetEnd      (seeq_t *);
-char      * seeqGetString   (seeq_t *);
-char      * seeqPrintError  (void);
+seeq_t     * seeqNew         (const char *, int);
+void         seeqFree        (seeq_t *);
+seeqfile_t * seeqOpen        (const char *);
+int          seeqClose       (seeqfile_t *);
+long         seeqFileMatch   (seeqfile_t *, seeq_t *, int);
+long         seeqStringMatch (const char *, seeq_t *, int);
+size_t       seeqGetLine     (seeq_t *);
+size_t       seeqGetStart    (seeq_t *);
+size_t       seeqGetEnd      (seeq_t *);
+int          seeqGetDistance (seeq_t *);
+char       * seeqGetString   (seeq_t *);
+const char * seeqPrintError  (void);
 
 #define RESET       "\033[0m"
 #define BOLDRED     "\033[1m\033[31m"      /* Bold Red */

### 2015-05-08 src/seeq-main.c

@@ -30,7 +30,12 @@
 #include <stdio.h>
 #include <getopt.h>
 
-char *USAGE = "Usage:"
+void say_usage(void);
+void say_version(void);
+void say_help(void);
+void SIGSEGV_handler(int) __attribute__ ((noreturn));
+
+static const char *USAGE = "Usage:"
 "  seeq [options] pattern inputfile\n"
 "    -v --version         print version\n"
 "    -z --verbose         verbose using stderr\n"
@@ -47,14 +52,13 @@ char *USAGE = "Usage:"
 "    -r --prefix          print only the prefix, ending before the match\n"
 "    -b --best            scan the whole line to find the best match (default: first match only)\n";
 
-int  seeq (char * expression, char * input, struct seeqarg_t args);
 void say_usage(void) { fprintf(stderr, "%s\n", USAGE); }
 void say_version(void) { fprintf(stderr, SEEQ_VERSION "\n"); }
 void say_help(void) { fprintf(stderr, "use '-h' for help.\n"); }
 
 void SIGSEGV_handler(int sig) {
    void *array[10];
-   size_t size;
+   int size;
 
    // get void*'s for all entries on the stack
    size = backtrace(array, 10);
@@ -301,6 +305,8 @@ main
          say_version();
          say_usage();
          exit(0);
+
+      default:
          break;
       }
    }

### 2015-05-08 src/seeq.c

@@ -87,6 +87,7 @@ seeq
    seeqfile_t * sqfile = seeqOpen(input);
    if (sqfile == NULL) {
       fprintf(stderr, "error in 'seeqOpen()': %s\n", seeqPrintError());
+      seeqFree(sq);
       return EXIT_FAILURE;
    }
 
@@ -107,12 +108,12 @@ seeq
       else match_options = SQ_MATCH;
       if (args.best) match_options |= SQ_BEST;
 
-      int retval;
+      long retval;
       while ((retval = seeqFileMatch(sqfile, sq, match_options)) > 0) {
-         if (args.compact) fprintf(stdout, "%ld:%d-%d:%d",sq->match.line, sq->match.start, sq->match.end-1, sq->match.dist);
+         if (args.compact) fprintf(stdout, "%ld:%ld-%ld:%d",sq->match.line, sq->match.start, sq->match.end-1, sq->match.dist);
          else {
             if (args.showline) fprintf(stdout, "%ld ", sq->match.line);
-            if (args.showpos)  fprintf(stdout, "%d-%d ", sq->match.start, sq->match.end-1);
+            if (args.showpos)  fprintf(stdout, "%ld-%ld ", sq->match.start, sq->match.end-1);
             if (args.showdist) fprintf(stdout, "%d ", sq->match.dist);
             if (args.matchonly) {
                sq->match.string[sq->match.end] = 0;

### 2015-05-08 src/seeqcore.h

@@ -47,57 +47,66 @@ typedef struct edge_t   edge_t;
 typedef struct trie_t   trie_t;
 typedef struct node_t   node_t;
 
-typedef unsigned int uint;
-
 struct node_t {
-   uint child[TRIE_CHILDREN];
-   uint parent;
+   size_t child[TRIE_CHILDREN];
+   size_t parent;
 };
 
 struct trie_t {
-   uint    pos;
-   uint    size;
-   uint    height;
+   size_t  pos;
+   size_t  size;
+   size_t  height;
    node_t  nodes[];
 };
 
 struct edge_t {
-   uint state;
-   int  match;
-   int  min_to_match;
+   size_t       state;
+   int          match;
+   unsigned int min_to_match;
 };
 
 struct vertex_t {
-   uint    node_id;
+   size_t  node_id;
    edge_t  next[NBASES];
 };
 
 struct dfa_t {
-   uint      pos;
-   uint      size;
-   trie_t  * trie;
-   vertex_t  states[];
+   size_t     pos;
+   size_t     size;
+   trie_t   * trie;
+   vertex_t   states[];
 };
 
 
+
+//   [0 ... 255] = 6,
+//   ['a'] = 0, ['c'] = 1, ['g'] = 2, ['t'] = 3, ['u'] = 3, ['n'] = 4, ['\0'] = 5,
+//   ['A'] = 0, ['C'] = 1, ['G'] = 2, ['T'] = 3, ['U'] = 3, ['N'] = 4, ['\n'] = 5
 static const int translate[256] = {
-   [0 ... 255] = 6,
-   ['a'] = 0, ['c'] = 1, ['g'] = 2, ['t'] = 3, ['u'] = 3, ['n'] = 4, ['\0'] = 5,
-   ['A'] = 0, ['C'] = 1, ['G'] = 2, ['T'] = 3, ['U'] = 3, ['N'] = 4, ['\n'] = 5
+   5,6,6,6,6,6,6,6,6,6, 5,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6,
+   6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6,
[…]

### 2015-05-08 test/testset.c

@@ -157,24 +157,20 @@ test_trie_new
    }
    }
 
-   for (int i = 0 ; i > -100 ; i--) {
-   for (int j = 0 ; j > -100 ; j--) {
-      trie_t *trie = trie_new(i*100, j*100);
-      g_assert(trie != NULL);
-      g_assert_cmpint(trie->pos, ==, 1);
-      g_assert_cmpint(trie->size, ==, 1);
-      g_assert_cmpint(trie->height, ==, 0);
-      g_assert_cmpint(trie->nodes[0].parent, ==, 0);
-      for (int k = 0; k < TRIE_CHILDREN; k++)
+   trie_t *trie = trie_new(0, 0);
+   g_assert(trie != NULL);
+   g_assert_cmpint(trie->pos, ==, 1);
+   g_assert_cmpint(trie->size, ==, 1);
+   g_assert_cmpint(trie->height, ==, 0);
+   g_assert_cmpint(trie->nodes[0].parent, ==, 0);
+   for (int k = 0; k < TRIE_CHILDREN; k++)
       g_assert_cmpint(trie->nodes[0].child[k], ==, 0);
-      free(trie);
-   }
-   }
+   free(trie);
 
    // Alloc test.
    mute_stderr();
    set_alloc_failure_rate_to(1.1);
-   trie_t * ret = trie_new(1, 10);
+   trie_t *ret = trie_new(1, 10);
    reset_alloc();
    g_assert(ret == NULL);
 
@@ -368,22 +364,21 @@ test_trie_search_getrow
 
    char path[10] = {0};
    // Insert random paths and search them.
-   srand48(123);
    for (int i = 0 ; i < 10000 ; i++) {
       for (int j = 0 ; j < 10 ; j++) {
          path[j] = (lrand48() % TRIE_CHILDREN);
       }
-      uint value = (uint) lrand48();
-      uint dfastate = (uint) lrand48();
-      uint retval, dfaval;
-      uint id = trie_insert(&trie, path, value, dfastate);
-      uint * row = trie_getrow(trie, id);
+      size_t value = (size_t) lrand48();
+      size_t dfastate = (size_t) lrand48();
+      size_t retval, dfaval;
+      size_t id = trie_insert(&trie, path, value, dfastate);
+      int * row = trie_getrow(trie, id);
 
[…]

### 2015-05-08 src/libseeq.c
Added 'skip' option.

@@ -340,6 +340,11 @@ seeqStringMatch
 //             * SQ_FIRST: searches the line until the first match.
 //             * SQ_BEST: searches the whole line to find the best match, i.e. the one with
 //               minimum matching distance. In case of many best matches, the first is returned.
+//
+//             * SQ_ILLEGAL_STOP: stops searching the current line if an illegal character is
+//               found (allowed characters are 'a','A','c','C','g','G','t','T','u','U','n','N',
+//               '\0','\n').
+//             * SQ_ILLEGAL_CONT: illegal characters will be substituted by mismatches ('N').
 //             
 //             If SQ_COUNTLINES or SQ_COUNTMATCH are specified, the other options are ignored.
 //             Also, the match/line count is passed as the return value.
@@ -369,6 +374,9 @@ seeqStringMatch
    // Count replaces all other options.
    int countall   = options & SQ_COUNTMATCH;
    int best_match = options & SQ_BEST;
+   const int * translate;
+   if (options & SQ_CONT) translate = translate_n;
+   else translate = translate_halt;
    if (options & (SQ_COUNTLINES|SQ_COUNTMATCH)) options = 0;
    else if ((options & 0x03) == 0) options = SQ_MATCH | SQ_NOMATCH;
 

### 2015-05-08 src/libseeq.h

@@ -40,6 +40,8 @@
 #define SQ_COUNTMATCH 0x08
 #define SQ_BEST       0x10
 #define SQ_FIRST      0x00
+#define SQ_CONT       0x20
+#define SQ_SKIP       0x00
 
 #include <stdio.h>
 

### 2015-05-08 src/seeq-main.c

@@ -46,11 +46,12 @@ static const char *USAGE = "Usage:"
 "    -n --no-printline    does not print the matching line\n"
 "    -l --lines           prints the original line number of the match\n"
 "    -p --positions       prints the position of the match within the matched line\n"
-"    -s --print-dist      prints the Levenshtein distance wrt the pattern\n"
+"    -k --print-dist      prints the Levenshtein distance wrt the pattern\n"
 "    -f --compact         prints output in compact format (line:pos:dist)\n"
 "    -e --end             print only the end of the line, starting after the match\n"
 "    -r --prefix          print only the prefix, ending before the match\n"
-"    -b --best            scan the whole line to find the best match (default: first match only)\n";
+"    -b --best            scan the whole line to find the best match (default: first match only)\n"
+"    -s --skip            skip lines containing non-DNA characters\n";
 
 void say_usage(void) { fprintf(stderr, "%s\n", USAGE); }
 void say_version(void) { fprintf(stderr, SEEQ_VERSION "\n"); }
@@ -94,6 +95,7 @@ main
    int endline_flag   = -1;
    int prefix_flag    = -1;
    int best_flag      = -1;
+   int skip_flag      = -1;
 
    // Unset options (value 'UNSET').
    input = NULL;
@@ -111,7 +113,8 @@ main
          {"positions",     no_argument, 0, 'p'},
          {"match-only",    no_argument, 0, 'm'},
          {"no-printline",  no_argument, 0, 'n'},
-         {"print-dist",    no_argument, 0, 's'},
+         {"print-dist",    no_argument, 0, 'k'},
+         {"skip",          no_argument, 0, 's'},
          {"lines",         no_argument, 0, 'l'},
          {"count",         no_argument, 0, 'c'},
          {"invert",        no_argument, 0, 'i'},
@@ -126,7 +129,7 @@ main
          {0, 0, 0, 0}
       };
 
-      c = getopt_long(argc, argv, "pmnislczfvherbd:",
+      c = getopt_long(argc, argv, "pmnislczfvkherbd:",
             long_options, &option_index);
  
       /* Detect the end of the options. */
@@ -179,6 +182,19 @@ main
             return EXIT_FAILURE;
          }
          break;
+
+      case 's':
+         if (skip_flag < 0) {
+            skip_flag = 1;
+         }
+         else {
+            say_version();
[…]

### 2015-05-08 src/seeq.c

@@ -107,6 +107,7 @@ seeq
       if (args.invert) match_options = SQ_NOMATCH;
       else match_options = SQ_MATCH;
       if (args.best) match_options |= SQ_BEST;
+      if (!args.skip) match_options |= SQ_CONT;
 
       long retval;
       while ((retval = seeqFileMatch(sqfile, sq, match_options)) > 0) {

### 2015-05-08 src/seeq.h

@@ -17,6 +17,7 @@ struct seeqarg_t {
    int prefix;
    int invert;
    int best;
+   int skip;
 };
 
 int  seeq (char * expression, char * input, struct seeqarg_t args);

### 2015-05-08 src/seeqcore.h

@@ -82,7 +82,7 @@ struct dfa_t {
 //   [0 ... 255] = 6,
 //   ['a'] = 0, ['c'] = 1, ['g'] = 2, ['t'] = 3, ['u'] = 3, ['n'] = 4, ['\0'] = 5,
 //   ['A'] = 0, ['C'] = 1, ['G'] = 2, ['T'] = 3, ['U'] = 3, ['N'] = 4, ['\n'] = 5
-static const int translate[256] = {
+static const int translate_halt[256] = {
    5,6,6,6,6,6,6,6,6,6, 5,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6,
    6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6,
    6,6,6,6,6,0,6,1,6,6, 6,2,6,6,6,6,6,6,4,6, 6,6,6,6,3,3,6,6,6,6,
@@ -93,6 +93,19 @@ static const int translate[256] = {
    6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6,6,6,6,6,
    6,6,6,6,6,6,6,6,6,6, 6,6,6,6,6,6
 };
+
+static const int translate_n[256] = {
+   5,4,4,4,4,4,4,4,4,4, 5,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4,
+   4,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4,
+   4,4,4,4,4,0,4,1,4,4, 4,2,4,4,4,4,4,4,4,4, 4,4,4,4,3,3,4,4,4,4,
+   4,4,4,4,4,4,4,0,4,1, 4,4,4,2,4,4,4,4,4,4, 4,4,4,4,4,4,3,3,4,4,
+   4,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4,
+   4,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4, 4,4,4,4,4,4,4,4,4,4,
[…]

### 2015-05-11 doc/response.tex
Updated response.

@@
 compared to 54 seconds for Linux \texttt{wc}, which only counts the
-number of characters. On such problems where the number of operations
-per character is small, the function used to read the file makes
-the difference. We used GNU's \texttt{readline} because it was
-faster than the alternatives in our initial tests.
+number of characters, words and lines. On such problems where the
+number of operations per character is small, the function used to read
+the file makes the difference. We use GNU's \texttt{readline} because
+it was faster than the alternatives in our tests.
 
@@
 due to cache misses or to the additional time required to process
-the hits (of which there is a lot in this example).
+the hits (of which there is a lot in this example), or both.
 
@@
 In comparison, Myers' bitfield algorithm ran for 66 minutes
-44 seconds on this problem. Counting 17 operations per character,
-this is about 75 ns per operation, which is the right order of
-magnitude. For the bitfield algorithm to beat Seeq on the last problem,
-each operation would have to be processed in less than 2.3 ns. We
-were very surprised by this result, because Myers' bitfield
+44 seconds on this problem. %Counting 17 operations per character,
+%this is about 75 ns per operation, which is the right order of
+%magnitude. For the bitfield algorithm to beat Seeq on the last problem,
+%each operation would have to be processed in less than 2.3 ns.
+We were very surprised by this result, because Myers' bitfield
 algorithm has the reputation of being the fastest known alignment
@@
 most of the time Seeq does just 1, and it is often a cache hit. On
-the last problem, Seeq performs 1 operation on 99.7\% of the
+the last example, Seeq performs 1 operation on 99.7\% of the
 nucleotides. In summary, Myers' bitfield is not the fastest
@@
 every job, it is not intended to replace BLAST or BWA. Yet there
-is a sweet spot for pattens around 20-30 nucleotides where we
-reach good performance even for a hig number of tolerated errors.
+is a sweet spot for pattens of tens of nucleotides where we
+reach good performance even for a big number of tolerated errors.
 
-Most sequences of this size are unique, even in large genomes,
+Most sequences of 20-30 nt are unique, even in large genomes,
 which is why spike-ins, watermarks, adapters, barcodes and

### 2015-05-11 test/testset.c
Updated test cases.

@@ -584,47 +584,47 @@ test_dfa_step
 
    edge_t transition;   
    // text[0]
-   g_assert(0 == dfa_step(0, translate[(int)text[0]], plen, tau, &dfa, exp, &transition));
+   g_assert(0 == dfa_step(0, translate_halt[(int)text[0]], plen, tau, &dfa, exp, &transition));
    g_assert_cmpint(transition.state, ==, 1);
    g_assert_cmpint(transition.match, ==, 2);
    // text[1]
-   g_assert(0 == dfa_step(transition.state, translate[(int)text[1]], plen, tau, &dfa, exp, &transition));
+   g_assert(0 == dfa_step(transition.state, translate_halt[(int)text[1]], plen, tau, &dfa, exp, &transition));
    g_assert_cmpint(transition.state, ==, 2);
    g_assert_cmpint(transition.match, ==, 2);
    // text[2]
-   g_assert(0 == dfa_step(transition.state, translate[(int)text[2]], plen, tau, &dfa, exp, &transition));
+   g_assert(0 == dfa_step(transition.state, translate_halt[(int)text[2]], plen, tau, &dfa, exp, &transition));
    g_assert_cmpint(transition.state, ==, 3);
    g_assert_cmpint(transition.match, ==, 2);
    // text[3]
-   g_assert(0 == dfa_step(transition.state, translate[(int)text[3]], plen, tau, &dfa, exp, &transition));
+   g_assert(0 == dfa_step(transition.state, translate_halt[(int)text[3]], plen, tau, &dfa, exp, &transition));
    g_assert_cmpint(transition.state, ==, 3);
    g_assert_cmpint(transition.match, ==, 2);
    // text[4]
-   g_assert(0 == dfa_step(transition.state, translate[(int)text[4]], plen, tau, &dfa, exp, &transition));
+   g_assert(0 == dfa_step(transition.state, translate_halt[(int)text[4]], plen, tau, &dfa, exp, &transition));
    g_assert_cmpint(transition.state, ==, 4);
    g_assert_cmpint(transition.match, ==, 2);
    // text[5]
-   g_assert(0 == dfa_step(transition.state, translate[(int)text[5]], plen, tau, &dfa, exp, &transition));
+   g_assert(0 == dfa_step(transition.state, translate_halt[(int)text[5]], plen, tau, &dfa, exp, &transition));
[…]

### 2015-05-11 src/seeq.c
Added memory report to verbose.

@@ -152,6 +152,11 @@ seeq
    }
    
    if (verbose) {
+      size_t * data = (size_t *) sq->dfa;
+      size_t mem_dfa  = *data * 16;
+      size_t mem_trie = *(size_t *)(*(data + 2)) * 32;
+      double mb = 1024.0*1024.0;
+      fprintf(stderr, "memory: %.2f MB (DFA: %.2f MB, trie: %.2f MB)\n", (mem_dfa + mem_trie)/mb, mem_dfa/mb, mem_trie/mb);
       fprintf(stderr, "done in %.3fs\n", (clock()-clk)*1.0/CLOCKS_PER_SEC);
    }
    

### 2015-05-12 src/libseeq.c
Increased memory efficiency with flexible trie.

@@ -738,6 +738,11 @@ dfa_new
       return NULL;
    }
 
+   // Fill struct.
+   dfa->size = vertices;
+   dfa->pos  = 1;
+   dfa->trie = trie;
+
    // Compute initial NFA state.
    char * path = malloc((size_t)wlen+1);
    if (path == NULL) return NULL;
@@ -745,7 +750,7 @@ dfa_new
    for (int i = tau + 1; i < wlen; i++) path[i] = 1;
 
    // Insert initial state into trie.
-   size_t nodeid = trie_insert(&trie, path, (size_t)tau+1, 0);
+   size_t nodeid = trie_insert(dfa, path, (size_t)tau + 1, (size_t)tau+1, 0);
    free(path);
 
    if (nodeid == (size_t)-1) {
@@ -753,11 +758,6 @@ dfa_new
       free(dfa);
       return NULL;
    }
-
-   // Fill struct.
-   dfa->size = vertices;
-   dfa->pos  = 1;
-   dfa->trie = trie;
 
    // Link trie leaf with DFA vertex.
    dfa->states[0].node_id = nodeid;
@@ -845,11 +845,19 @@ dfa_step
       // If exists, just link with the existing state.
       dfa->states[dfa_state].next[base].state = dfalink;
    } else if (exists == 0) {
-      if (dfa_newstate(dfap, code, prev, base, dfa_state) == -1) return -1;
+      if (dfa_newstate(dfap, code, prev, base, dfa_state) == -1) {
+         free(state);
+         free(code);
+         return -1;
+      }
       // Set min_to_match:
       dfa = *dfap;
       dfa->states[dfa_state].next[base].min_to_match = (unsigned int)(plen - last_active);
-   } else return -1;
+   } else {
+      free(state);
+      free(code);
+      return -1;
+   }
 
    free(state);
    free(code);
@@ -936,8 +944,14 @@ dfa_newstate
    // Set error to 0.
    seeqerr = 0;
 
+   trie_t ** triep = &((*dfap)->trie);
+
+   // Compute length of the constant tail.
+   size_t minlen = (*triep)->height;
+   while (code[minlen-1] == 1) minlen--;
+
    // Insert new NFA state in the trie.
-   size_t nodeid = trie_insert(&((*dfap)->trie), code, (size_t)alignval, (*dfap)->pos);
+   size_t nodeid = trie_insert(*dfap, code, minlen, (size_t)alignval, (*dfap)->pos);
[…]

### 2015-05-12 src/seeq.c

@@ -153,7 +153,7 @@ seeq
    
    if (verbose) {
       size_t * data = (size_t *) sq->dfa;
-      size_t mem_dfa  = *data * 16;
+      size_t mem_dfa  = *data * (5*16+8);
       size_t mem_trie = *(size_t *)(*(data + 2)) * 32;
       double mb = 1024.0*1024.0;
       fprintf(stderr, "memory: %.2f MB (DFA: %.2f MB, trie: %.2f MB)\n", (mem_dfa + mem_trie)/mb, mem_dfa/mb, mem_trie/mb);

### 2015-05-12 src/seeqcore.h

@@ -118,8 +118,9 @@ int         dfa_step      (size_t, int, int, int, dfa_t **, char *, edge_t *);
 void        dfa_free      (dfa_t *);
 trie_t    * trie_new      (size_t, size_t);
 int         trie_search   (trie_t *, char*, size_t*, size_t*);
-size_t      trie_insert   (trie_t **, char*, size_t, size_t);
+size_t      trie_insert   (dfa_t *, char*, size_t, size_t, size_t);
 int       * trie_getrow   (trie_t *, size_t);
+size_t      trie_newnode  (trie_t **, size_t);
 void        trie_reset    (trie_t *);
 
 #define RESET       "\033[0m"

### 2015-05-12 src/seeq.c
Added number of states in memory summary.

@@ -156,7 +156,7 @@ seeq
       size_t mem_dfa  = *data * (5*16+8);
       size_t mem_trie = *(size_t *)(*(data + 2)) * 32;
       double mb = 1024.0*1024.0;
-      fprintf(stderr, "memory: %.2f MB (DFA: %.2f MB, trie: %.2f MB)\n", (mem_dfa + mem_trie)/mb, mem_dfa/mb, mem_trie/mb);
+      fprintf(stderr, "memory: %.2f MB (DFA: %ld in %.2f MB, trie: %ld in %.2f MB)\n", (mem_dfa + mem_trie)/mb, *data, mem_dfa/mb, *(size_t *)(*(data + 2)), mem_trie/mb);
       fprintf(stderr, "done in %.3fs\n", (clock()-clk)*1.0/CLOCKS_PER_SEC);
    }
    

### 2015-06-04 src/libseeq.c
Corrected major bug in all matches mode.

@@ -277,8 +277,7 @@ seeqStringMatch
       // Match accept condition:
       // If there is overlap with the previous, accept only if new_dist < last_dist.
       // 
-      if (streak_dist <= sq->tau && streak_dist <= current_dist && !match && (i >= overlap || streak_dist < last_d) && (!opt_best || streak_dist < best_d)) {
-         if (streak_dist < current_dist) match = 1;
+      if (streak_dist < current_dist && !match && (i >= overlap || streak_dist < last_d) && (!opt_best || streak_dist < best_d)) {
          size_t j = 0;
          uint32_t rnode = 1;
          int d = sq->tau + 1;
@@ -300,10 +299,11 @@ seeqStringMatch
             // Check whether we can delete some nucleotides at the start to avoid
             // the overlap.
             size_t gap = last_end - match_start;
-            if (last_end - match_start <= (size_t) (d - sq->tau)) {
+            if (last_end - match_start <= (size_t) (sq->tau - d)) {
                match_start += gap;
                match_dist  += gap;
             } else {
+               streak_dist = current_dist;
                match = 0;
                continue;
             }
@@ -322,6 +322,7 @@ seeqStringMatch
          last_end = match_end;
          last_d   = match_dist;
          overlap  = i + (size_t)(sq->wlen - sq->tau);
+         match = 1;
          // Break if done.
          if (!all_match) end = 1;
       }

### 2015-06-04 doc/response.tex

--- 2015-05-07 doc/response.tex
@@
 editor and of our peers, but we do not consider that we should
-interrupt a fertile discussion because the manuscript was
+interrupt a fruitful discussion because the manuscript was
 rejected. Please do not see in this letter a criticism of the
-reviews, but simply a wish to respond to some of the points that
-were raised.
+reviews, but simply a wish to touch upon questions that were
+impossible to treat in a two page manuscript and to respond to
+some of the points that were raised.
 
-Because this is not a rebuttal, we will do our best to make it
-informal and didactic.
-But before doing so, we would like to congratulate the reviewers for
-a really outstanding work. We are all for software quality and
-excellent science, we wish that every tool were scrutinized with
-the same amout of care. We have also learned important lessons,
-and the suggestions will greatly help us improve the software and
-the manuscript.
+Since this is not a rebuttal, we will do our best to make it
+informal and didactic. But before doing so, we would like to
+thank the reviewers for a really outstanding work. We are
+all for software quality and excellent science, we wish that every
+tool were scrutinized with the same amout of care. We have learned
+important lessons, and the suggestions will greatly help us improve
+the software and the manuscript.
 
@@
 
-It started from casual Internet
-browsing of pages that have little to do with bioinformatics.
-The first is a seminal series of blog posts by Cox Russel [1]
+Our work on Seeq started with casual Internet browsing.
+The first source was a seminal series of blog posts by Russ Cox [1]
 explaining how and why he implemented the regular expression library
@@
 (including those of Perl and Python) have ``forgotten'' the
-theory and are clearly dominated by lazy DFA-based implementations.
+theory and many developers are unaware that lazy DFA-based
+implementations outperform the alternatives.
 
@@
 approximate string query. In Fig.~\ref{generalities} we give some
-generalities about Levenshtein automata, in case the reader is not
-familiar with these concepts.
+generalities about Levenshtein automata, we also refer the interested
+reader to the blog post of Nick Johnson, which is very clear and makes
+a lively introduction to the subject, and to reference [3] for a
+more comprehensive treatment.
 
@@
 are labelled by two numbers indicating their number of errors and
-their position relative to the search pattern. NFAs can be ins
+their position relative to the search pattern. NFAs can be in
 several active states (dark circles). The initial active states
-of the NFA are as shown in the second panel. After reading
+of the NFA are as shown on the second panel. After reading
 \texttt{A} from the input, the active states are updated as shown
@@
 
-When we had to identify spike oligos in our
-sequencing experiments, we used the Python implementation suggested
+When we had to identify oligos that we spiked in our sequencing
+experiments for quantification purposes,
+we gave a try to the Python implementation suggested
 by Nick Johnson, which solved the problem (Illumina sequencing is
@@
 the Levenshtein distance). However, the Python prototype was rather
-slow, and because it is a non lazy DFA, the memory footprint was high.
+slow, and because it is a non lazy DFA, the memory footprint was too
+high to allow more than two errors.
 
-We connected the ideas of the two posts, and the first C
-implementation of Seeq gave astonishing results. We knew that this
-was not novel (lazy DFAs were implemented in egrep since 1989), but
-because Seeq was much faster then Myers' bitfield algorithm,
-we thought that lazy Levenshtein automata did not quite
-make the jump from string processing to bioinformatics and that
-someone should tell the community that they are efficient.
+We connected the ideas of the two posts and we used the trick of
+reversing the DFA to identify the position of the hit in the string
+as suggested by Russ Cox. The first C implementation of Seeq gave
+very good results and we decided to continue. We came across the
+literature on lazy/partial DFAs later, but we knew from the outset
+that the ideas were rather old  (for the record, lazy DFAs were
+introduced by Alfred Aho in egrep in 1989).
+
+So we agree with reviewer 2 that the algorithm is not novel, yet
+we are curious where s/he finds any mention of lazy DFAs in
+references [4-6] and where the extensive analysis is performed in
+[3]. The numerous questions that s/he raises about the performance
+and the behavior of the algorithm rather point to the conclusion
+that little is know about this class of algorithms.
+
+To our knowledge, most of the theoretical work on lazy DFAs is
+due to Gonzalo Navarro [7] and was performed in the context of
+English language. Importantly, the corpus consisted of 10 MB texts,
+the significance of which we will explain below. We do not know for
+sure why this line of research
+was discontinued but the following quote from [7] may give a hint
+``Memory limitations prevented us to compute automata with
+more than 500,000 states (although we used a 128 Mb machine for
+this task)''. That's right, even on a 128 Mb machine...
+Myer's BPM algorithm (which has a nearly invisible
+memory footprint) [4] outcompeted most alignment methods on the
+hardware of that decade and quickly gained the reputation of being
+the fastest known alignment algorithm. But hardware evolves, and
+problems too. As reviewer 2 points out, it would be interesting
+to see how lazy DFAs compare to Myer's BPM on modern hardware.
+But before doing so, we need to deal with a more pressing issue.
 
 \section{Sparsity of the DFA}
-\label{sketch}
 
-At that stage, we were both surprised and skeptical. One concern (also
+Our initial prototype was fast, but our main concern (also
 pointed out by reviewer 2) was that the potential number of states
-was very high. Our initial implementation was based on the NFA
-called the edit graph (Fig.~\ref{generalities}a),
-which can be in $2^{(k+1)L}$ states, where $k$
-is the tolerated number of errors and $L$ is the length of the
-pattern. This number is enormous, but surprisingly, the memory
-footprint was never an issue,
-even for large files and long patterns. For some reason,
-the lazy Levenshtein automaton explored only a small portion of its
-states.
+was very high. The NFA representation gives the intuition that the
+DFA may have close to $3^L$ states, where $L$ is the
+length of the pattern. For a modest 25 nucleotide pattern, this
+is over 800 billion states and there is no way to store that many.
+But there is also no way to reach that many states,
+because this would take an input file of at least 800 Gb. The
+question is not how many states the DFA can have, but how many
+states the input will generate. Literature has little to offer
+about this question. Once again we refer to the comment of Gonzalo
+Navarro [7] that ``this analysis is pessimistic and the results
+are much better in practice, as well as the efficiency ratio among
+partial and full DFA''.
 
-We realized that we could reduce the possible number of states by
-keeping only the upper envelope of the NFA states in the edit graph
-(in the manuscript, we presented states as columns from the alignment
-because we thought that bioinformaticians would be more comfortable
-with this representation, but the representations are strictly
-equivalent). This made Seeq faster and reduced the memory footprint.
-And it also allows us to explain why the DFA is sparse.
-
-With this representation, a state is a line of the edit graph that
-extends from the top-left
-corner. On this line, local maxima have travelled from the top-left
-corner by going either to the right (in case of a match) or in
-diagonal (in case of a mismatch). So the probability that a local
-maximum is at position $(i,j)$, where $(0,0)$ is
-the top-left corner is approximately
-
-\begin{equation}
-{j \choose i} \left(\frac{3}{4}\right)^i
-  \left(\frac{1}{4}\right)^{j-i}.
-\end{equation}
-
-This expresses that the number of diagonal steps has a binomial
-distribution with probability $3/4$. For fixed $j$, the average
-depth of local maxima is $3j/4$. Using R, we can get the 95\%
-percentiles of $i$ for the first 20 values of $j$. We obtain
-\texttt{0  0  1  1  2  3  3  4  5  5  6  6  7  8  8  9 10 10 11 12}.
-All the DFA states are lines above the diagonal, and approximately
-95\% of them will be beneath the 95\% percentiles of $i$. A graphical
-representation for a pattern of 20 nucleotides with up to 12 errors
-gives the idea that the majority of the states are squeezed within
-a narrow region of the edit graph (Fig.~\ref{sparsity}).
+Why would this be the case? In the edit graph representation
+(Fig.~\ref{generalities}a), a state of the DFA is a line of connected
+active states that extends from the top-left corner. As a heuristic
+argument, consider that active sates are independent and move forward
+in case of a match (with probabiliy 1/4), and in diagonal in case of
+a mismatch (with probability 3/4). The height
+of an active node then has a binomial distribution, of which we can
+compute the 95-th quantiles for each column. For a pattern of
+25 nucleotides we obtain \texttt{0  0  1  1  2  3  3  4  5  5  6  6
+7  8  8  9 10 10 11 12 12 13 14 14 15}. This gives an area of the
+edit graph that will contain the states visited most frequently,
+which we have plotted in Fig.~\ref{sparsity} for a tolerance of 
+12 errors.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[scale=0.24]{sparsity_of_the_DFA.pdf}}
+\centerline{\includegraphics[scale=0.65]{sparsity_of_the_DFA.pdf}}
 \caption{The Levenshtein automaton of Seeq is sparse. The sketch
-represents the edit graph for a pattern of length 20 with 12
+represents the edit graph for a pattern of length 25 with 12
 tolerated errors. The dark dots cover a region where active
-states are approximately 90-95\% of the time. This region is
+states are approximately 95\% of the time. This region is
 substantially smaller than the full edit graph and contains
@@
 
-It is not trivial to compute the number of states in this region
-in general terms, however it can be done for the example above.
-There are 221,930 states \textit{i.e.} only $0.006\%$ of the
-$3.5 \cdot 10^9$ upper bound computed from the formula provided by
-reviewer 2. This is of course a heuristic argument that does not
-resist worst case analysis. But the upshot is that on average,
-Seeq needs to store very few states of the Levenshtein automaton.
+Computing the number of states in this region in general terms is
+not trivial, however in the example above there are 221,930 states
+\textit{i.e.} less than one millionth of the upper bound $3^{25}$.
+In reality, active states are not independent, so this sketch
+only has a heuristic value. The upshot is that the DFA will spend
+most of its time in a small number of states, and the probability
+that other states are visited is low. Biological sequences are
+not random streams, but their structured and repeated nature will
+reduce the number of visited states rather than increase it.
 
@@
 
-The sketch of analysis above explains all the properties of Seeq.
-Because the number of states is much lower than expected, Seeq
-rarely needs to use dynamic programming to compute a new state.
-As an added benefit, there will be few cache misses because
-most of the time, Seeq only references the same small
-set of sates.
+The key (and underappreciated) strength of Seeq is that the number
+of states visited by a Levenshtein automaton is much lower than
+expected. The immediate consequences are that the memory won't blow
+up and that you will rarely use dynamic programming. Most of the
+time, Seeq just follows pointers, and because lazy DFAs have good
+locality [1], many of those are cache hits.
 
-Let us give a few examples to highlight the behavior of the
-algorithm. When searching the 20 nucleotide pattern
-\texttt{GATGTAGCGCGATTAGCCTG}
-in hg19 with up to 4 errors, Seeq ended with 5,294 states and spent
-0.01\% of the time in dynamic programming. Even though
-it had to scan the whole genome, the run was executed in 27 seconds,
-compared to 54 seconds for Linux \texttt{wc}, which only counts the
-number of characters. On such problems where the number of operations
-per character is small, the function used to read the file makes
-the difference. We used GNU's \texttt{readline} because it was
-faster than the alternatives in our initial tests.
+Upon reading the reviewers comments, we realized that the benchmark
+shown in the manuscript is not convincing. Our bad. We hope that
+we can do better in this response in order to give a feel for the
+performance of Seeq. Reviewer 2 suggested to benchmark Seeq against
+Myer's bitvector algorithm, SeqAn and Razers3. SeqAn is a very high
+quality library, but it takes some time to get familiar with, so we
+are still not proficient to write competitive software with it.
 
-Allowing 12 errors, which is the case of the sketch
-of proof given in section \ref{sketch}, Seeq ended with 170,679 states,
-close to our approximation. In this case, the running time
-was 1 minute 18 seconds, of which 0.39 seconds were spent in dynamic
-programming. So the difference with the previous case is either
-due to cache misses or to the additional time required to process
-the hits (of which there is a lot in this example).
+In what follows we search approximate matches of sequences of length 20,
+27, 34 and 42 in the humgan genome. All the sequences are prefixes of
+the random sequence \texttt{GATGTAGCGCGATTAGCCTGAAAATGCGAGTACGGCGCGAAT}.
+The input is a fasta file of hg19 (sequences were downloaded from [8],
+concatenated and every chromosome was put on a single line). The
+software was run on a 16-core dual-processor Intel Xeon with 256 GB
+of DDR3-RAM at 1,866 MHz and a solid state hard drive.
 
-What about longer patterns with higher error tolerance?
-When searching the 42 nucleotide pattern
-\texttt{GATGTAGCGCGATTAGCCTGAAAATGCGAGTACGGCGCGAAT} with
-up to 12 errors, Seeq ran for 2 minutes 11 seconds and ended with
-3,486,701 states. About 20 seconds were spent in dynamic programming,
-which shows that it is not the bottleneck.
-Even with very efficient vectorization to speed up this step, the
-performance of Seeq would not improve substantially. The increase
-in running time is more likely due to the increased number of cache
-misses.
+The running time of the naive implementation of Myer's bitvector
+algorithm (\textit{i.e.} as provided in [4]) ran between 47 and 50
+minutes. As expected from the conditions of our benchmark, the running
+time is nearly constant. However, this is far from competitive, as
+can be seen in Table~\ref{benchmark}. The running time of Razers3
+ranged from 53 seconds to ~6 minutes, and that of Seeq from 10 seconds
+to ~3 minutes. On many of these problems, Seeq ran faster than
+the Linux \texttt{wc} command (30 seconds), which only counts
+characters. Seeq is about five times faster than Razers3 in full
+sensitivity mode and two to five times faster than Razers3 in 99\%
+sensitivity mode.
 
-In comparison, Myers' bitfield algorithm ran for 66 minutes
-44 seconds on this problem. Counting 17 operations per character,
-this is about 75 ns per operation, which is the right order of
-magnitude. For the bitfield algorithm to beat Seeq on the last problem,
-each operation would have to be processed in less than 2.3 ns. We
-were very surprised by this result, because Myers' bitfield
-algorithm has the reputation of being the fastest known alignment
-algorithm. But it needs to do 17 operations per character, whereas
-most of the time Seeq does just 1, and it is often a cache hit. On
-the last problem, Seeq performs 1 operation on 99.7\% of the
-nucleotides. In summary, Myers' bitfield is not the fastest
-alignment algorithm, it is the fastest alignment algorithm with
-no memory footprint.
+\begin{table}[h]
+\centering
+\begin{tabular}{c|c||c|c|c|c||c|c}
+  \multicolumn{2}{c}{} & \multicolumn{4}{c}{Seeq} &
+      \multicolumn{2}{c}{Razers3} \\
+  \hline
+  m  & k & time (s) & \# hits & mem (MB) & DP (s) & time (s) & \# hits \\
+  \hline
+  20 & 3  & 10  & 58     & 0.11    & 0.00 & 53/53 & 58/58 \\
+  20 & 4  & 10  & 2908   & 0.46    & 0.00 & 76/77 & 2908/2908 \\
+  20 & 5  & 10  & 68566  & 1.45    & 0.03 & 82/106 & 63803/68647 \\
+  27 & 3  & 10  & 0      & 0.10    & 0.00 & 61/61 & 0/0 \\
+  27 & 4  & 10  & 0      & 0.32    & 0.02 & 64/69 & 0/0 \\
+  27 & 5  & 10  & 29     & 0.95    & 0.04 & 59/96 & 23/29 \\
+  27 & 6  & 11  & 552    & 2.70    & 0.06 & 99/199 & 550/552 \\
+  27 & 7  & 14  & 8795   & 7.76    & 0.19 & 98/201 & 8622/8796 \\
+  27 & 8  & 17  & 107809 & 21.9    & 0.54 & -/205 & -/107909 \\
+  34 & 5  & 10  & 0      & 0.94    & 0.02 & 63/67 & 0/0 \\
+  34 & 6  & 11  & 0      & 2.48    & 0.10 & 79/108 & 0/0 \\
+  34 & 7  & 13  & 1      & 5.83    & 0.24 & 68/102 & 1/1 \\
+  34 & 8  & 19  & 37     & 13.10   & 0.50 & 120/247 & 37/37 \\
+  34 & 9  & 23  & 527    & 28.22   & 1.04 & 120/256 & 524/527 \\
+  34 & 10 & 29  & 6797   & 50.51   & 2.26 & -/252 & -/6798 \\
+  42 & 8  & 17  & 0      & 13.92   & 0.57 & 66/73 & 0/0 \\
+  42 & 9  & 23  & 0      & 29.91   & 1.34 & 65/123 & 0/0 \\
+  42 & 10 & 30  & 1      & 61.20   & 2.39 & 122/340 & 1/1 \\
+  42 & 11 & 39  & 6      & 121.47  & 5.14 & 126/344 & 6/6 \\
+  42 & 12 & 53  & 52     & 232.71  & 9.30 & 123/347 & 52/52 \\
+  42 & 13 & 75  & 772    & 424.35  & 18.06 & -/352 & -/772 \\
+  42 & 14 & 109 & 8664   & 738.23  & 32.17 & 46/46 & 0/0 \\
+  42 & 15 & 161 & 82507  & 1260.50 & 55.44 & 46/46 & 0/0 \\
+  42 & 15 & 177 & 82507  & 555.91* & 71.58 \\
+\end{tabular}
+\caption{Benchmark of Seeq against Razers3. We searched patterns of
+length 20 to 42 (parameter m) with errors ranging from 3 to 15
+(parameter k) in the human genome (a fasta file of hg19). For Seeq
+we recorded the running time, the number of hits discovered, the
+memory usage of the DFA and the time spend in dynamic programming.
+For Razers3, we recorded the running time and the number of hits
+discovered. The first number was obtained by using the default sensitivity
+\texttt{-rr 99} and the second by using the maximum sensitivity 
+\texttt{-rr 100}. In the last case, Seeq was run with the option
+\texttt{-y500} to limit the memory usage to 500 MB, which is why
+the memory usage is marked by a star. Note that Seeq does not
+allow overlapping matches while Razers3 does, which explains that
+the number of hits can be different. A dash (-) indicates that
+Razers3 crashed with segmentation fault.
+\label{benchmark}
+}
+\end{table}
 
-If lazy Levenshtein automata are so good, why were they not used
-widely before? This is a good question. We surmise that this has
-to do with the evolution of hardware. At the time they were proposed
-in bioinformatics, \textit{i.e.} at the end of the 1990's, computers
-typically had around 100 Mb of RAM, so they were probably considered
-memory hungry and other methods filled the niche.
+We also recorded the memory usage of Seeq. The most difficult problem
+requested over 1.2 GB of RAM, but limiting the memory to ~500 MB
+increased the running time by only 9\%. Also notice that the
+memory footprint did not exceed 22 MB for patterns of 27
+nucleotides, which is an empirical confirmation of the heuristic
+argument above that the number of visited states is insignificant
+compared to the worst-case estimates.
 
-It is important to note that in the last case, the memory footprint
-peaked at 1.5 Gb, which is approximately 430 bytes per state. The
-states of the DFA proper are only an array of 5 pointers (40 bytes),
-but Seeq indexes them in a ternary tree, which represents the bulk
-of the memory usage. So it is true that the number of states will
-become intractable for very large patterns and very high number of
-tolerated errors.
+The last indicator that we collected is the time spent in dyamic
+programming. Except for the most difficult problems, this number
+does not exceed 10\% of the total running time, which means that
+in many cases, the user could not notice any increase in speed even
+if we made this step instantaneous. It is important to realize that
+even on the most difficult problems, dynamic programming is performed
+very few times. To give a figure, this happens once every 368
+nucleotides in the last problem of Table~\ref{benchmark} (and once
+every 1.6 million nucleotides in the first). Only when searching
+long patterns with high error rates would it pay off to use
+modern hardware to accelerate dynamic programming. But long patterns
+with high error rates is the turf of other algorithms (for instance
+BLAST).
 
-\section{Concluding remarks}
+We also noticed that Razers3 crashed on the most difficult problems
+or even failed silently. We will get in contact with the maintainers
+about this issue. Also, we would like to point out that Razers3 is
+a mapper, so performing the benchmark with only one sequence
+at a time is not showing its true potential. We used it in this
+condition to better highlight the differences of the algorithms
+running in the background, but we do not claim that Seeq is a point
+blank replacement of Razers3.
 
-We still believe that Seeq has an interesting message for the
-bioinformatics community. It is certainly not the right tool for
-every job, it is not intended to replace BLAST or BWA. Yet there
-is a sweet spot for pattens around 20-30 nucleotides where we
-reach good performance even for a hig number of tolerated errors.
+\begin{figure}[!tpb]
+\centerline{\includegraphics[scale=.90]{run_course.pdf}}
+\caption{Processing course of a Seeq run. The line represents the
+time required to process the given amount of input on the last
+problem shown in Table~\ref{benchmark} (without memory limit).
+Notice the run is slowest at the beginning and reaches its steady
+state after ~10\% of the input has been processed.
+}\label{course}
+\end{figure}
 
-Most sequences of this size are unique, even in large genomes,
-which is why spike-ins, watermarks, adapters, barcodes and
-other artificial sequences that need to be identified in sequencing
-experiments will be in this range. The target application of 
-Seeq is thus the Python library because it allows to easily write
-parsers to identify the pattern and further process the reads as
-simple Python strings.
+Finally, we wish to highlight a key property of Seeq and lazy DFAs.
+Fig.~\ref{course} shows a processing course on the last
+problem of Table~\ref{benchmark} (without memory limit). As expected,
+the run is slow at the beginning and only achieves a steady state
+speed after ~10\% of the input has been processed, which is about
+300 MB. Remember that lazy DFAs were initially benchmarked on much
+smaller files, which explains that their performance was only
+slightly better, if not worse than other algorithms.
 
-We are aware that the information provided here would have made
-the manuscript much stronger, but with a two page limit we had to
-remove important parts. The difficulty was that Seeq is not novel
-enough to deserve a full feature article, and it is too complex
-to be described in just two pages.
-We decided to present Seeq from the
-perspective of the user, explain the principle and show the
-performance. To be honest, we thought that there would be a
-round of revisions where we could explain all this and decide
-together with the reviewers how much should go in the article.
-We realize now that we did not do a very good job at explaining
-why Seeq is interesting. But we learned our lessons, we will do
-better next time.
+\section{Conclusion}
+
+Lazy DFAs have been analyzed at a time when they had mostly a
+theoretical interest. They can be implemented efficiently on
+modern hardware where their advantage over other algorithms
+increases with file size. The core engine of Seeq is not novel,
+it is actually an antique. And yet its performance is unmatched
+for the reasons explained above.
+
+We believe that this message is important for the bioinformatics
+community. Not only because Seeq is useful, but because
+lazy DFAs will have applications in many problems.
+
+\section{Now what?}
+
+Retrospectively, we agree that the manuscript was weak and rejecting
+it was the right decision. Because of the two-page limit, we opted
+to write it from the perspective of the user and missed major points
+about the relevance of lazy DFAs for modern problems. Yet, we
+believe that Seeq is useful and that it carries a simple and important
+message for modern bioinformatics. This is why we ask the permission
+to submit another version of the manuscript. Here is a summary of
+that actions that we may take at this stage.
+
+\vspace{.2in}
+
+\noindent
+\textbf{What we will do}.
+\noindent
+Regardless of what happens to this submission, we will
+\begin{enumerate}
+\item Write a complete documentation in pdf format.
+\item Improve the code as per the comments of reviewer 1.
+\item Rewrite the manuscript completely.
+\item Change the benchmark as indicated by reviewer 2.
+\end{enumerate}
+
+\noindent
+\textbf{What we will not do.}
+\noindent
+For the reasons explained above, we will not
+\begin{enumerate}
+\item Use vectorization to speed up dynamic programming.
+\end{enumerate}
+
+\noindent
+\textbf{What we may do.}
+\noindent
+Depending on the decision about the manuscript and how
+much the following is justified, we may
+\begin{enumerate}
+\item Write a full rebuttal. We will of course write a
+formal rebuttal if this work is considered for
+publication.
+\item Paralellize the algorithm. This is possible only when
+searching multiple patterns, which is not the core job of
+Seeq.
+\item Implement tunable costs and affine gaps. In this
+case the threshold distance is less meaningful for the
+user, but we can do it if some applications require it.
+\end{enumerate}
 
@@
 [2] http://blog.notdot.net/2010/07/Damn-Cool-Algorithms-Levenshtein-Automata
+\par\noindent
+[3] A guided tour to approximate string matching. Navarro, Gonzalo.
+ACM computing surveys (CSUR), 2001
+\par\noindent
+[4] A fast bit-vector algorithm for approximate string matching based
+on dynamic programming. G Myers - Journal of the ACM (JACM), 1999
+\par\noindent
+[5] Finding approximate patterns in strings. E Ukkonen - J. Algorithms,
+1985.
+\par\noindent
+[6] Approximate string matching by finite automata. Melichar, Bořivoj.
+Computer Analysis of Images and Patterns. Springer Berlin Heidelberg, 1995.
+\par\noindent
+[7] Approximate text searching. PhD thesis, Depat. of Computer Siences,
+University of Chile, December 1998. pp. 118-124.
+\par\noindent
+[8]http://hgdownload.cse.ucsc.edu/goldenpath/hg19/chromosomes/
 

### 2015-06-04 src/libseeq.c
Corrected bug in matching algorithm.

@@ -263,11 +263,11 @@ seeqStringMatch
       else if (cin == 6 && stream_opt) continue;
       else if (cin == 7 && opt_ignore) continue;
       else if (cin >= 5) {
-         current_dist = sq->tau+1;
+         current_dist = sq->tau;
          end = 1;
       }
       if (slen - i - 1 < (size_t) min_to_match) {
-         current_dist = sq->tau + 1;
+         current_dist = sq->tau;
          end = 1;
       }
 
@@ -277,7 +277,7 @@ seeqStringMatch
       // Match accept condition:
       // If there is overlap with the previous, accept only if new_dist < last_dist.
       // 
-      if (streak_dist < current_dist && !match && (i >= overlap || streak_dist < last_d) && (!opt_best || streak_dist < best_d)) {
+      if (streak_dist <= sq->tau && streak_dist <= current_dist && !match && (i >= overlap || streak_dist < last_d) && (!opt_best || streak_dist < best_d)) {
          size_t j = 0;
          uint32_t rnode = 1;
          int d = sq->tau + 1;

### 2015-06-04 test/testset.c

@@ -1019,6 +1019,7 @@ test_seeqFileMatch
 
    // Realloc match stack.
    sq = seeqNew("A", 0, 0);
+   seeqStringMatch("AAAAAAAAAAAAAAAAAAAA", sq, SQ_ALL);
    g_assert_cmpint(seeqStringMatch("AAAAAAAAAAAAAAAAAAAA", sq, SQ_ALL), == , 20);
    for (int i = 0; i < 20; i++) {
       match = seeqMatchIter(sq);

### 2015-06-04 src/libseeq.c
Final code cleaning and test set repaired.

@@ -904,8 +904,7 @@ dfa_newstate
    seeqerr = 0;
 
    // Check memory usage.
-   size_t memory = (*dfap)->pos * sizeof(vertex_t); // DFA memory.
-   memory += (*dfap)->pos * ((*dfap)->trie->height/5 + ((*dfap)->trie->height%5 > 0)); // Cached alignments.
+   size_t memory = (*dfap)->pos * (*dfap)->state_size; // DFA memory.
    memory += (*dfap)->trie->pos * sizeof(node_t); // Trie memory.
    if ((*dfap)->maxmemory > 0 && memory > (*dfap)->maxmemory) {
       (*dfap)->trie = realloc((*dfap)->trie, sizeof(trie_t) + (*dfap)->trie->pos * sizeof(node_t));

### 2015-06-04 test/testset.c

@@ -182,9 +182,8 @@ test_trie_insert
 
    // Insert path references.
    for (int i = 0; i < 8; i++) {
-      uint8_t * code = malloc(trie_height/5 + (trie_height%5 > 0));
-      path_encode(test_path[i], code, trie_height);
-      dfa->states[i+1].align = code;
+      vertex_t * vertex = (vertex_t *) (dfa->states + (i+1) * dfa->state_size);
+      path_encode(test_path[i], vertex->code, trie_height);
    }
 
    // Trie contains path 2,1,1,1,1,1...
@@ -309,9 +308,8 @@ test_trie_insert
 
    // Insert path references.
    for (int i = 0; i < 3; i++) {
-      uint8_t * code = malloc(trie_height/5 + (trie_height%5 > 0));
-      path_encode(test_path2[i], code, trie_height);
-      lowdfa->states[i+1].align = code;
+      vertex_t * vertex = (vertex_t *) (lowdfa->states + (i+1) * lowdfa->state_size);
+      path_encode(test_path2[i], vertex->code, trie_height);
    }
 
 
@@ -372,9 +370,8 @@ test_trie_search
 
    // Insert path references.
    for (int i = 0; i < 7; i++) {
-      uint8_t * code = malloc(trie_height/5 + (trie_height%5 > 0));
-      path_encode(test_path[i], code, trie_height);
-      dfa->states[i+2].align = code;
+      vertex_t * vertex = (vertex_t *) (dfa->states + (i+2) * dfa->state_size);
+      path_encode(test_path[i], vertex->code, trie_height);
       g_assert(trie_insert(dfa, test_path[i], i+2) == 0);
    }
 
@@ -473,10 +470,10 @@ test_dfa_newvertex
    for (int i = 1; i < 1023; i++) {
       uint32_t new = dfa_newvertex(&dfa);
       g_assert_cmpint(dfa->pos, ==, i+2);
-      g_assert(dfa->states[new].align == NULL);
-      g_assert_cmpint(dfa->states[new].match, ==, DFA_COMPUTE);
+      vertex_t * vertex = (vertex_t *) (dfa->states + new * dfa->state_size);
+      g_assert_cmpint(vertex->match, ==, DFA_COMPUTE);
       for (int j = 0; j < NBASES; j++) {
[…]
